From Parametricity to Conservation Laws, via Noether's Theorem

### 2013-07-11 mechanics/paper.tex
Introduction, and Section 2

  m\ddot{x_2} = k(x_1 - x_2)
\end{mathpar}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them.

The Lagrangian (\ref{eq:coupled-spring-lagrangian}) does not refer to
any fixed point in space; only the relative distance between the two
particles, along with their velocities, is relevant. Therefore, the
Lagrangian (\ref{eq:coupled-spring-lagrangian}) is invariant under
translation in space by some arbitrary displacement $y$:
\begin{equation}\label{eq:coupled-spring-translation-invariance}
  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = L(t,x_1+y,x_2+y,\dot{x_1},\dot{x_2})
\end{equation}
By Noether's theorem, invariance under spatial translation implies
that the linear momentum of the whole system is constant for all
time. For this system, conservation of linear momentum is stated
mathematically as a vanishing derivative with respect to time:
\begin{equation}\label{eq:coupled-spring-linear-momentum}
  \frac{d}{\mathit{dt}}~m(\dot{x}_1 + \dot{x_2}) = 0
\end{equation}
In general, Noether's theorem gives us a way of deriving conserved
properties like (\ref{eq:coupled-spring-linear-momentum}) from
invariance properties like
(\ref{eq:coupled-spring-translation-invariance}). In this case, we
have used invariance under translation in space to derive conservation
of linear momentum. Other common examples include invariance under
translation in time, yielding conservation of energy, and invariance
under rotation, yielding conservation of angular momentum. We will see
examples of each of these kinds of invariance, and their consequent
conservation laws, in Section~\ref{sec:examples}.

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminiscent of the free theorem we stated for the polymorphic program
at the start of this introduction. FIXME

In this paper, we present a way of using a generalised version of
Reynolds' theory of relational parametricity to prove the geometric
invariance properties required for the hypotheses of Noether's
theorem. We build an extension of System F$\omega$ suitable for
writing invariant Lagrangians. In our system, the Lagrangian
(\ref{eq:coupled-spring-lagrangian}) we gave above for describing a
system of two particles coupled by a spring will have the type:
\begin{displaymath}
  \forall x \mathord: \mathsf{T}(1).~C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, x \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{displaymath}
where... FIXME

Atkey, Johann and Kennedy \cite{atkey13abstraction} have already
presented a polymorphic type system for expressing geometric
invariance properties similar to the

FIXME: area of triangle type

By extending System F$\omega$, we get to

@@

Our major contributions are threefold:
\begin{enumerate}
\item FIXME: reformulate AJK in terms of the reflexive graph model of
  relational parametricity in order to unify that paper with the
  natural extension of Reynolds' relational parametricity to indexed
  types.
\item FIXME: Present an extension of System F$\omega$ suitable for
  writing invariant Lagrangians. This requires special attention to
  ensure that all the functions that we can write are actually
  continuous and differentiable. Hence we include a special type of
  smooth, invariant functions between cartesian spaces.
\item FIXME: Present many examples of 
\end{enumerate}

@@

\begin{itemize}
\item FIXME: Recall the necessary background for Noether's theorem
  (Section~\ref{sec:conservation-laws-from-symmetry}).
\item FIXME: Show how to get symmetry properties from types
  (Section~\ref{sec:symmetry-from-types}).
\item FIXME: In Section~\ref{sec:types-for-classical-mech}, we
  construct a type system for classical mechanics,
\item FIXME: Section~\ref{sec:examples} we present examples of using
  our extension of System F$\omega$ for writing invariant Lagrangians
  that describe physical systems.
\item FIXME: field theory
\end{itemize}

@@

Make sure to state up front the necessary background required. To
understand: basic(ish) calculus. We will be using the Calculus of
Variations, but only by stating the main results. We go for standard
physics notation, rather than anything more complicated.

@@

Introduce by stating Newton's second law, and saying that it can be
derived from the principle of stationary action.

Analytical mechanics, Lagrangian and Hamiltonian mechanics. Basically
different mathematical reformulations of Newtonian mechanics that
yield new insights into the underlying structure of the kinds of
physical systems studied in classical mechanics.

@@

Lagrangian mechanics reformulates Newtonian mechanics in terms of

Two points: the Lagrangian and generalised coordinates. Do
generalised coordinates as a secondary thing.

\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}

@@

\begin{example}
  In Example~\ref{ex:invariance-examples}, we noted that the
  Lagrangians (\ref{eq:coupled-spring-lagrangian}) and
  (\ref{eq:simple-g-lagrangian}) are both invariant under arbitrary
  translations in time. In terms of continuous invariance, their
  corresponding actions are invariant under the families
  $\Phi_\epsilon(t) = t + \epsilon$ and $\Psi_\epsilon(q) =
  q$. Plugging these definitions into
  Equation~(\ref{eq:conservation-law}), we learn that, for these
  systems:
  \begin{displaymath}
    \frac{d}{dt}~\left(L - \sum_{i=1}^2 \dot{q_i}\frac{\partial L}{\partial \dot{q_i}} \right) = 0
  \end{displaymath}
  For the spring-coupled particles Lagrangian,
  (\ref{eq:coupled-spring-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2) \right) = 0
  \end{displaymath}
  and for the single particle under gravity Lagrangian
  (\ref{eq:simple-g-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x}^2 + \dot{y}^2) + mgy\right) = 0
  \end{displaymath}
  In both cases, we have discovered that, as a consequence of
  invariance under translation in time, the total energy of the system
  is conserved (note that the $-$ signs in the original Lagrangians
  have turned into $+$s). In general, invariance under translation in
  time implies, via Noether's theorem, conservation of energy.
\end{example}

This finishes our short introduction to Lagrangian mechanics and
Noether's theorem. FIXME: State that we will do more examples later
on, after we discuss how to derive invariance properties from types.

@@

\subsection{A Surface Syntax for Smooth Functions}

@@

\section{Basic Field Theory}

\subsection{Principle of Stationary Action for Fields}

\subsection{Noether's Theorem for Fields}

\subsection{The Klein-Gordon Equation}

\section{Conclusions}

>>>>

  m\ddot{x_2} = k(x_1 - x_2)
\end{mathpar}
We could now proceed to solve these ODEs to further analyse the
behaviour of this system. However, Noether's theorem gives us a
powerful way of gaining insight into properties of these ODEs ``for
free'', without necessarily having to find solutions to them. This is
accomplished by finding invariance properties of the Lagrangian.

The Lagrangian (\ref{eq:coupled-spring-lagrangian}) does not refer to
any fixed point in space; only the relative distance between the two
particles, along with their velocities, is relevant. Therefore, the
Lagrangian (\ref{eq:coupled-spring-lagrangian}) is invariant under
translation in space by some arbitrary displacement $y$:
\begin{equation}\label{eq:coupled-spring-translation-invariance}
  L(t,x_1,x_2,\dot{x_1},\dot{x_2}) = L(t,x_1+y,x_2+y,\dot{x_1},\dot{x_2})
\end{equation}
By Noether's theorem, invariance under spatial translation implies
that the linear momentum of the whole system is constant for all
time. For this system, conservation of linear momentum is stated
mathematically as the vanishing of the derivative with respect to time
of the total linear momentum of the system:
\begin{equation}\label{eq:coupled-spring-linear-momentum}
  \frac{d}{\mathit{dt}}~m(\dot{x}_1 + \dot{x_2}) = 0
\end{equation}
In general, Noether's theorem gives us a way of deriving conserved
properties like (\ref{eq:coupled-spring-linear-momentum}) from
invariance properties like
(\ref{eq:coupled-spring-translation-invariance}). In this case, we
have used invariance under translation in space to derive conservation
of linear momentum. Other common examples include invariance under
translation in time, yielding conservation of energy, and invariance
under rotation, yielding conservation of angular momentum. We will see
examples of each of these kinds of invariance, and their consequent
conservation laws, in Section~\ref{sec:examples}.

The invariance property stated in
Equation~(\ref{eq:coupled-spring-translation-invariance}) is highly
reminiscent of the free theorem we stated for the polymorphic program
at the start of this introduction. A change in representation in the
input, whether a change in data representation from a type $\alpha$ to
a type $\beta$, or a change in where time zero is counted from,
results in no change in the output. Since Reynolds' relational
parametricity was allowed us to derive the invariance properties for
the program $f$ just by looking at its type, we ask the following
question:

\medskip

\noindent
\textbf{Question:} \emph{Is it possible to use Reynolds' theory of
  relational parametricity to derive the invariance properties of
  Lagrangians required for Noether's theorem to derive conservation
  laws?}

\medskip

In this paper, we answer this question positively. We show that it is
indeed the case that we can use a generalised version of Reynolds'
theory of relational parametricity to prove the geometric invariance
properties required for the hypotheses of Noether's theorem. We
construct an extension of System F$\omega$ suitable for writing
invariant Lagrangians, with a relationally parametric model that
allows us to prove invariance properties as free theorems. In our
system, the Lagrangian (\ref{eq:coupled-spring-lagrangian}) we gave
above for describing a system of two particles coupled by a spring
will have the type:
\begin{displaymath}
  \begin{array}{@{}l}
  \forall y \mathord: \mathsf{T}(1).\\
  \quad C^\infty(\mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, y \rangle \times \mathbb{R}\langle 1, 0 \rangle \times \mathbb{R}\langle 1, 0 \rangle, \mathbb{R}\langle 1, 0 \rangle)
\end{array}
\end{displaymath}
We explain this type by breaking it down from left to right. The
quantifier $\forall x \mathord: \mathsf{T}(1)$ indicates that we are
quantifying over all translations $y$ in one-dimensional space, just
as we did implicity in Equation
(\ref{eq:coupled-spring-translation-invariance}). The notation
$C^\infty(-,-)$ denotes the type of \emph{smooth} functions between
spaces. A smooth function is a function that is infinitely
differentiable. All the functions that we wish to use as Lagrangians
must be smooth (at least up to second order) for the theory to
work. The type $\mathbb{R}\langle g, x \rangle$ denotes real numbers
that vary with some linear transformation $g$ and translation $x$. In
this case, all the linear transformations are $1$, indicating no
transformation. Likewise, when the translation component is $0$, there
is no translation. Thus a value of type $\mathbb{R}\langle 1, 0
\rangle$ is invariant, but a value of type $\mathbb{R}\langle 1, y
\rangle$ varies with the translation $y$. In the smooth function type
above, in domain space consists of vectors of five real numbers: the
time, the two particle positions, which vary with the translation $y$,
and the two particle velocities.

Atkey, Johann and Kennedy \cite{atkey13abstraction}, building on older
work by Kennedy \cite{kennedy94para}, have already presented a
polymorphic type system for expressing geometric invariance properties
similar to the translation invariance property in Equation
(\ref{eq:coupled-spring-translation-invariance}). In the present work,
we extend their type system by embedding it within System
F$\omega$. The benefits of this embedding are twofold. Firstly, by
using the richer type-level structure of System F$\omega$, we can
easily add useful indexed types like length indexed vectors and smooth
function types, both of which would have required special treatment in
the special purpose type system presented by Atkey \emph{et
  al.}. Secondly, by using System F$\omega$ as our base, we are able
to reuse the reflexive graph-based relationally parametric semantics
presented by Atkey \cite{atkey12relational}, extending it as
appropriate for our geometric setting.

@@

Our major contributions are threefold:
\begin{enumerate}
\item We reformulate the type system for geometric invariance of
  Atkey, Johann and Kennedy \cite{atkey13abstraction} as an extension
  of System F$\omega$. This shows that the special status of the
  group-indexed types in Atkey \emph{et al.}'s work can be
  incorporated into a standard framework for presenting indexed types.
\item We present a relationally parametric semantics of our extension
  of System F$\omega$. In this model, following Atkey's model for
  basic System F$\omega$ \cite{atkey12relational}, each kind is
  interpreted as a reflexive graph. Our key technical contribution is
  to note that the many geometric groups that we wish to incorporate
  are expressible as groupoids, and hence as reflexive graphs. We
  therefore show that Reynolds' theory of relational parametricity,
  suitably generalised to indexed types by considering reflexive
  graphs, already accomodates geometric invariance properties.
\item Finally, we connect the free theorems derivable from our
  relationally parametric semantics with Noether's theorem for
  deriving conservation laws. We present many examples of invariant
  Lagrangians expressible in our type system, along with the
  consequent conservation laws for physical systems. In each case, the
  necessary invariance property is derived from the type we assign to
  the Lagrangian.
\end{enumerate}

@@

\begin{itemize}
\item In the next section,
  Section~\ref{sec:conservation-laws-from-symmetry}, we describe the
  necessary background to understand Noether's theorem and the
  conservation laws it generates. We briefly introduce the Lagrangian
  formulation of classical mechanics, define precisely what it means
  for a physical system to be invariant under transformations, and
  state Noether's theorem.
\item Having shown how invariance generates conservation laws in
  Section~\ref{sec:conservation-laws-from-symmetry}, in
  Section~\ref{sec:symmetry-from-types}, we show how to derive
  invariance properties from indexed types. We present a relationally
  parametric semantics for System F$\omega$, and show how geometric
  invariance properties can be accomodated within this semantics.
\item In Section~\ref{sec:types-for-classical-mech}, we construct a
  type system for classical mechanics, by extending System F$\omega$
  with the necessary constants for smooth function types and
  combinators for constructing smooth functions. Programming directly
  with our combinators in raw System F$\omega$ is awkward, due to the
  point-free style required, so in Section~\ref{sec:surface-syntax} we
  define a surface syntax for defining smooth functions that is
  translated into our applied System F$\omega$.
\item In Section~\ref{sec:examples} we present several examples of
  using our extension of System F$\omega$ for writing invariant
  Lagrangians that describe physical systems.
\item Section~\ref{sec:conclusions} concludes, and offers directions
  for further work.
\end{itemize}

@@

Noether's theorem applies to systems described using the formalism of
Lagrangian mechanics, a reformulation of Newtonian classical
mechanics. In this section we can only offer a very brief introduction
to the concepts of Lagrangian mechanics. The reader is referred to
other sources, such as Landau and Lifschitz \cite{landau60mechanics}
or Arnol'd \cite{arnoldXXmathematical} for the necessary background in
the mathematical theory of classical mechanics. Lagrangian mechanics,
and Noether's theorem, make crucial use of the Calculus of
Variations. Gelfand and Fomin \cite{gelfandXXcalculus} provide a good
introduction to this topic.

In this section, we will endeavour to keep the presentation relatively
elementary. However, some knowledge of differential and integral
calculus will be required. Knowledge of basic Newtonian mechanics will
also be helpful. We will use standard notation from physics and
mathematics for integrals and (partial) derivatives. In particular,
when discussing derivatives with respect to time, we use dot notation:
$\dot{x}$ for the first derivative of $x$ with respect to time, and
$\ddot{x}$ for the second derivative.

@@

Lagrangian mechanics is a branch of analytical mechanics, along with
Hamiltonian mechanics
\cite{landau60mechanics,arnoldXXmathematical}. Analytical mechanics
seeks to discover the underlying structure of Newtonian classical
mechanics by reformulating it in different terms.

@@

For Lagrangian mechanics, classical mechanical systems are described
in terms of \emph{Lagrangians} $L = T - V$, where $T$ is the total
kinetic energy of the system, and $V$ is the total potential
energy. The Lagrangian is a function of time, the positions of all the
particles in the system, and their velocities. This is usually written
like so, with abuse of notation for $\dot{q}$ which is a variable
here, not a derivative of a function:
\begin{displaymath}
  L(t, q, \dot{q}) = T - V
\end{displaymath}
In this expression, $q$ and $\dot{q}$ are actually vectors containing
the components of the position and velocity vectors for each
particle. A key feature of Lagrangian mechanics is that the coordinate
system used to describe the system need not necessarily be the usual
cartesian coordinates. For example, one of the components of $q$ might
be the angle of swing of a pendulum. This use of generalised
coordinates is one of the advantages of Lagrangian mechanics over the
Newtonian presentation in terms of positions and forces.

The Lagrangian for a particular system is used to define the
\emph{action} of that system. For a particular path $q : \mathbb{R}
\to \mathbb{R}^n$ and endpoints $a < b$ in time, the action of the
Lagrangian $L$ over this path is defined by the following integral:
\begin{displaymath}
  \mathcal{S}[q;a;b] = \int_a^b L(t,q(t),\dot{q}(t)) \mathit{dt}
\end{displaymath}
Below, we will use the principle of stationary action to select the
physically realisable paths from all the possible paths $q :
\mathbb{R} \to \mathbb{R}^n$.

@@

\begin{example}
  In Example~\ref{ex:invariance-examples}, we noted that the
  Lagrangians (\ref{eq:coupled-spring-lagrangian}) and
  (\ref{eq:simple-g-lagrangian}) are both invariant under arbitrary
  translations in time. In terms of continuous invariance, their
  corresponding actions are invariant under the families
  $\Phi_\epsilon(t) = t + \epsilon$ and $\Psi_\epsilon(q) =
  q$. Plugging these definitions into
  Equation~(\ref{eq:conservation-law}), we learn that, for these
  systems:
  \begin{displaymath}
    \frac{d}{dt}~\left(L - \sum_{i=1}^2 \dot{q_i}\frac{\partial L}{\partial \dot{q_i}} \right) = 0
  \end{displaymath}
  For the spring-coupled particles Lagrangian,
  (\ref{eq:coupled-spring-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x_1}^2 + \dot{x_2}^2) + \frac{1}{2}k(x_1 - x_2)^2 \right) = 0
  \end{displaymath}
  and for the single particle under gravity Lagrangian
  (\ref{eq:simple-g-lagrangian}), we obtain:
  \begin{displaymath}
    \frac{d}{dt}~\left(\frac{1}{2}m(\dot{x}^2 + \dot{y}^2) + mgy\right) = 0
  \end{displaymath}
  In both cases, we have discovered that, as a consequence of
  invariance under translation in time, the total energy of the system
  is conserved (note that the $-$ signs in the original Lagrangians
  have turned into $+$s). In general, invariance under translation in
  time implies, via Noether's theorem, conservation of energy.
\end{example}

This finishes our short introduction to Lagrangian mechanics and
Noether's theorem. We have shown how Noether's theorem can be used to
derive conservation laws from invariance properties. We now go on to
show how suitable invariance properties can be derived from types. In
Section~\ref{sec:examples}, we give examples of invariant Lagrangian's
and their consequent conservation laws.

@@

\subsection{A Surface Syntax for Smooth Functions}
\label{sec:surface-syntax}

@@

% \section{Basic Field Theory}

% \subsection{Principle of Stationary Action for Fields}

% \subsection{Noether's Theorem for Fields}

% \subsection{The Klein-Gordon Equation}

\section{Conclusions}
\label{sec:conclusions}
